---
title: "1_1. exploratory analysis"
author: Dan Weinberger
date: "5/8/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(arrow) # for analyze parquet in R
library(tidyr) # for data analysis
library(dplyr) # for data analysis
library(ggplot2) # for plot
library(lubridate) # for creating date variable
library(stringr)
```

## Read in the parquet database

```{r} 
NY_pq <- open_dataset("R:/ParquetData/NY", format = "parquet") # a function in arrow package 
```

##Births

Extract the births to a data frame and save it
```{r, eval=F}
NY_births_ids <- NY_pq %>%
  rename(AMONTH='Admission month',AYEAR="Admission year"  ,BWT="Birth weight in grams", ZIP="Patient ZIP Code" ,ID="Visit linkage variable",birth_year="Birth year"  ) %>%
    filter(BWT!='') %>%
    dplyr::select(AMONTH, AYEAR, BWT, ZIP, ID,birth_year)%>%
  collect() %>%
  filter(AYEAR>=2016 & AYEAR<=2019 & !is.na(ID) & birth_year>=2016) #there are a bunch of people with a birth year (and age) that is large  

saveRDS(NY_births_ids, './CONFIDENTIAL/all_births.rds')
```


```{r}
NY_births_ids <- readRDS( './CONFIDENTIAL/all_births.rds') %>%
  mutate(BWT=as.numeric(BWT))
```

How many births per year?
```{r}
NY_births_N <- NY_births_ids %>%
  group_by(AYEAR) %>%
  filter(AYEAR>=2016 & AYEAR<=2019 ) %>%  
  summarize(N_cases = n()) %>%
  collect() %>%
  arrange(AYEAR)

NY_births_N
```


Birthweight is in grams. <2500 is low birthweight; <1500 is very low birthweight
```{r}
hist(NY_births_ids$BWT)
```
identify subsequent hospitalizations for these babies. 53,376 follow up visits for 844,461 births
```{r}
birth_ids <- NY_births_ids %>%pull(ID)

follow_up <- NY_pq %>%
  filter(`Birth year`>=2016) %>% #there are a bunch of people with birth year <<2016
    rename(AMONTH='Admission month',AYEAR="Admission year"  ,BWT="Birth weight in grams", ZIP="Patient ZIP Code" ,ID="Visit linkage variable",
           age_days='Age in days (when age < 1 year)',
           age_months="Age in months (when age < 11 years)" ,
           age_years = "Age in years at admission" ) %>%
filter(ID %in% birth_ids & BWT=='')%>% #identify visits for kids born in a hospital for post-birth visits
  dplyr::select(ID, ZIP, age_days,age_months,age_years,AMONTH,AYEAR, starts_with('ICD-10-CM Diagnosis'))%>%
  collect()%>%
  mutate(age_months= as.numeric(age_months),
         age_years = as.numeric(age_years),
         age_days=as.numeric(age_days),
    agem = if_else(age_years==0, age_days/30.437,age_months)) %>%
   rename_with(~paste0("ICD", str_extract(., "\\d+$")), starts_with("ICD-10-CM Diagnosis"))
```

Age distribution of visits
```{r}
hist(follow_up$agem)

```

follow_up

```{r}
follow_up %>% group_by(ID) %>% 
  summarize(N_visits=n()) %>%
  ungroup() %>%
  group_by(N_visits) %>%
  summarize(N_people=n())
```


```{r}
a1 <- follow_up %>%
  unite(all_icd, ICD1:ICD25, na.rm=F) %>%
  mutate( rsv = if_else(  grepl('B974', all_icd)| grepl('J121', all_icd)| grepl('J121', all_icd)| grepl('J205', all_icd)| grepl('J210', all_icd) , 1,0
                          )
          ) %>%
  arrange(ID, agem) %>%
  group_by(ID) %>%
  mutate(visitN=row_number()) %>%
  ungroup()
```

On which visit number did RSV occur?
```{r}
table(a1$rsv, a1$visitN)
```

```{r}
a1 %>%
  filter(rsv==1) %>%
  ggplot(aes(x=agem)) +
  geom_histogram()+
  theme_classic()+
  ggtitle('Age (m) of rsv cases admitted to hospital')
```


How many RSV visits does a person have?

```{r}
a1 %>% group_by(ID) %>% summarize(N_rsv=sum(rsv)) %>%
  ungroup()%>%
  group_by(N_rsv) %>%
  summarize(N_people=n())
```

ED data
"The SEDD capture emergency visits at hospital-owned emergency departments (EDs) that do not result in hospitalization. Information about patients initially seen in the ED and then admitted to the hospital is included in the State Inpatient Databases (SID). "
```{r}
birth_ids <- NY_births_ids %>%pull(ID)


NY_ed_pq <- open_dataset("R:/ParquetData/NY_SEDD", format = "parquet") %>% # a function in arrow package 
   rename(ID=VisitLink) %>%
filter(ID %in% birth_ids )%>% #identify visits for kids born in a hospital for post-birth visits
  dplyr::select(ID, ZIP, AGEDAY,AGEMONTH,AGE,AMONTH,AYEAR, starts_with('I10_DX1'))%>%
   dplyr::select(AGEMONTH, AGE,AGEDAY, AMONTH, AYEAR, ZIP,starts_with("I10_DX")) %>%    
  collect() %>%
    unite(all_icd, I10_DX1:I10_DX19, na.rm=F) %>%
  mutate( rsv = if_else(  grepl('B974', all_icd)| grepl('J121', all_icd)| grepl('J121', all_icd)| grepl('J205', all_icd)| grepl('J210', all_icd) , 1,0
                          )
          ) %>%
  filter(rsv==1)

saveRDS(NY_ed_pq, './CONFIDENTIAL/NY_ed_rsv.rds')


NY_ed_pq %>%
  mutate(age_months= as.numeric(AGEMONTH),
         age_years = as.numeric(AGE),
         age_days=as.numeric(AGEDAY),
    agem = if_else(age_years==0, age_days/30.437,age_months)) %>%
    ggplot(aes(x=agem)) +
  geom_histogram()+
  theme_classic()+
  ggtitle('Age (m) of rsv cases in the ED')
#8884 inpatient records, 6944 ed records

NY_ed_pq %>%
  mutate(age_months= as.numeric(AGEMONTH),
         age_years = as.numeric(AGE),
         age_days=as.numeric(AGEDAY),
    agem = if_else(age_years==0, age_days/30.437,age_months)) %>%
  summary(agem)
```

